A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-03-29 review/response.qmd

Below is a point-by-point explanation of how we have changed the paper. The reviewer comments are in black and our responses are in red. Reviewers' comments are in black and **\textcolor{purple}{our responses are in red}**.

# Reviewer: 1

I commend the authors for their work in the tidy index library.  Certainly, a user friendly and unified approach to indices is useful across many fields.  My main concern about the paper is that it doesn't seem to fit within the aims and scope of JCGS.  A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software.  This paper, while certainly useful, doesn't necessarily propose new statistical methods but rather proposes a great tidy framework for handling index calculation.  As such, for JCGS, I think the authors need to highlight the statistical contributions more than the library.  However, I don't think the authors statistical contributions (even if highlighted) are sufficiently novel to warrant publication in JCGS.  As written, I think the paper is more suited for publication in, say, Journal of Statistical Software than JCGS.

**\textcolor{purple}{we have highlighted the methodology contributions this framework makes in the conclusion: While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives. The pipeline proposed in the paper allows confidence interval to be calculated in the pipeline and sensitivity analysis to be incorporated in the index construction process.}**

### 2024-03-29 tidyindex.qmd
address reviewer 1 through emphasizing mothodology contribution in the conclusion

---
title: "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data"
format:
  tandf-pdf:
    keep-tex: true  
    linestretch: 2
pdf-engine: pdflatex
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: 
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-3813-7155
  - name: Ursula Laa
    affiliation: 
      - name: University of Natural Resources and Life Sciences
        department: Institute of Statistics
        address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
        city: Vienna
        country: Austria
    orcid: 0000-0002-0249-6439
  - name: Nicolas Langrené
    affiliation: 
      - name: BNU-HKBU United International College
        department: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science
        address: 2000 Jintong Road, Tangjiawan, Zhuhai
        city: Zhuhai, Guangdong
        country: China
    orcid: 0000-0001-7601-4618
  - name: Patricia Menéndez
    affiliation: 
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
      - name: University of Melbourne
        department: School of Mathematics and Statistics
        address: 813 Swanston Street, Parkville VIC 3052, Australia
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0003-0701-6315
abstract: |
  Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making. While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their values and variation in values. Here we discuss a modular data pipeline recommendation to assemble indexes. It is universally applicable to index computation and allows investigation of index behavior as part of the development procedure. One can compute indexes with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess indexes' robustness. The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on countries' ranking; and how to calculate bootstrap confidence intervals for the Standardized Precipitation Index (SPI). The methods are supported by a new R package, called `tidyindex`.
keywords:
  - indexes
  - data pipeline
  - software design
  - uncertainty
  - decision-making
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(width = 70)
library(tidyverse)
library(latex2exp)
library(patchwork)
library(tidyindex)
library(lmomco)
library(lubridate)
library(SPEI)
library(tsibble)
library(GGally)
library(patchwork)
library(tourr)
```

@@

Now, we introduce the notation used for describing pipeline modules. Consider a multivariate spatio-temporal process,

```{=tex}
\begin{equation}
\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}
```
where:

@@

Simplification takes a continuous variable $x_j(s;t)$ and categorises it into a discrete set $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$ through a piecewise constant function,

```{=tex}
\begin{equation}
v[x_i(s;t)] = 
\begin{cases}
a_0 & C_1 \leq x^i(s; t) < C_0 \\
a_1 & C_2 \leq x^i(s; t) < C_1 \\
a_2 & C_3 \leq x^i(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x^i(s; t)
\end{cases}
\end{equation}
```
This is typically used at the end of the index pipeline to simplify the index to communicate to the public the severity of the concept of interest measured by the index. An example of simplification is to map the calculated SPI to four categories: mild, moderate, severe, and extreme drought.

@@

The paper introduces a data pipeline comprising nine modules designed for the construction and analysis of indexes within the tidy framework. The pipeline offers a modular workflow to compute indexes with different parameterizations, to test minor changes to the original index definition, and to quantify uncertainties. The framework proposed in the paper is universal to index across diverse domains. Examples are provided, including the drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions, to illustrate how slight adjustment of linear combination weights impact the index, and to calculate confidence intervals on the index.

>>>>

---
title: "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data"
format:
  tandf-pdf:
    keep-tex: true  
    linestretch: 2
pdf-engine: pdflatex
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: University of Texas at Austin
        department: Department of Statistics and Data Science
        address: Welch 5.216, 105E 24th Street, Stop G3800
        city: Austin, Texas
        country: USA
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@austin.utexas.edu
  - name: Dianne Cook
    affiliation: 
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-3813-7155
  - name: Ursula Laa
    affiliation: 
      - name: University of Natural Resources and Life Sciences
        department: Institute of Statistics
        address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
        city: Vienna
        country: Austria
    orcid: 0000-0002-0249-6439
  - name: Nicolas Langrené
    affiliation: 
      - name: BNU-HKBU United International College
        department: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science
        address: 2000 Jintong Road, Tangjiawan, Zhuhai
        city: Zhuhai, Guangdong
        country: China
    orcid: 0000-0001-7601-4618
  - name: Patricia Menéndez
    affiliation: 
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
      - name: University of Melbourne
        department: School of Mathematics and Statistics
        address: 813 Swanston Street, Parkville VIC 3052, Australia
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0003-0701-6315
abstract: |
  Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making. While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their values and variation in values. Here we discuss a modular data pipeline recommendation to assemble indexes. It is universally applicable to index computation and allows investigation of index behavior as part of the development procedure. One can compute indexes with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess indexes' robustness. The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on countries' ranking; and how to calculate bootstrap confidence intervals for the Standardized Precipitation Index (SPI). The methods are supported by a new R package, called `tidyindex`.
keywords:
  - indexes
  - data pipeline
  - software design
  - uncertainty
  - decision-making
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(width = 70, pillow.print_max = 5, pillar.print_min = 5)
library(tidyverse)
library(latex2exp)
library(patchwork)
library(tidyindex)
library(lmomco)
library(lubridate)
library(SPEI)
library(tsibble)
library(GGally)
library(patchwork)
library(tourr)
```

@@

Now, we introduce the notation used for describing pipeline modules. Consider a multivariate spatio-temporal process, $$\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n $$
where:

@@

Simplification takes a continuous variable $x_j(s;t)$ and categorises it into a discrete set $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$ through a piecewise constant function,
```{=tex}
\begin{equation}
v[x_i(s;t)] = 
\begin{cases}
a_0 & C_1 \leq x^i(s; t) < C_0 \\
a_1 & C_2 \leq x^i(s; t) < C_1 \\
a_2 & C_3 \leq x^i(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x^i(s; t)
\end{cases}
\end{equation}
```
This is typically used at the end of the index pipeline to simplify the index to communicate to the public the severity of the concept of interest measured by the index. An example of simplification is to map the calculated SPI to four categories: mild, moderate, severe, and extreme drought.

@@

The paper introduces a data pipeline for constructing and analyzing indexes within the tidy framework. The pipeline comprises nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. The proposed pipeline is implemented as an R package, `tidyindex`, available on CRAN.

One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity analysis. While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives. The pipeline proposed in the paper allows for confidence interval calculation and sensitivity analysis to be incorporated in the index construction process, for indexes across various domains. These is illustrated through the three examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).

One future development of the work is to extend the tidyindex framework to support index calculation on commonly used domain-specific data formats, for example, NetCDF for climate indexes.

 <!-- Given the decisions analysts need to make when combining multivariate data into indexes, the proposed pipeline enables the evaluation of how the specific choice can affect the index, as well as how the index may appear under alternative options. Furthermore, uncertainty calculation can also flow through the pipeline, providing the index with confidence measures. -->

### 2024-03-29 tidyindex.tex
address reviewer 1 through emphasizing mothodology contribution in the conclusion

\title{A Tidy Framework and Infrastructure to Systematically Assemble
Spatio-temporal Indexes from Multivariate Data}
\author{H. Sherry
Zhang$\textsuperscript{1}$~\orcidlink{0000-0002-7122-1463}, Dianne
Cook$\textsuperscript{1}$~\orcidlink{0000-0002-3813-7155}, Ursula
Laa$\textsuperscript{2}$~\orcidlink{0000-0002-0249-6439}, Nicolas
Langrené$\textsuperscript{3}$~\orcidlink{0000-0001-7601-4618}, Patricia
Menéndez$\textsuperscript{1,4}$~\orcidlink{0000-0003-0701-6315}}

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{4} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

Now, we introduce the notation used for describing pipeline modules.
Consider a multivariate spatio-temporal process,

\begin{equation}
\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}

where:

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{DATA }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{module1}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
  \FunctionTok{module2}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
  \FunctionTok{module3}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
\NormalTok{  ...}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
# A tibble: 128,576 x 18
   .idx  .dist id         month       ym  prcp  tmax  tmin  tavg  long
   <chr> <chr> <chr>      <dbl>    <mth> <dbl> <dbl> <dbl> <dbl> <dbl>
 1 spei  gev   ASN000290~     6 1990 Jun   170  29.7  16.2  23.0  142.
 2 spei  gev   ASN000290~     7 1990 Jul   102  31.2  17.2  24.2  142.
 3 spei  gev   ASN000290~     8 1990 Aug     0  31.3  13.1  22.2  142.
 4 spei  gev   ASN000290~     9 1990 Sep     0  32.8  16.3  24.5  142.
 5 spei  gev   ASN000290~    10 1990 Oct     0  36.8  21.5  29.2  142.
 6 spei  gev   ASN000290~    11 1990 Nov   278  36.3  24.8  30.6  142.
 7 spei  gev   ASN000290~    12 1990 Dec  1869  34.4  24.5  29.4  142.
 8 spei  gev   ASN000290~    12 1990 Dec  1869  34.4  24.5  29.4  142.
 9 spei  gev   ASN000290~     1 1991 Jan  5088  31.2  24.4  27.8  142.
10 spei  gev   ASN000290~     1 1991 Jan  5088  31.2  24.4  27.8  142.
# i 128,566 more rows
# i 8 more variables: lat <dbl>, name <chr>, .pet <dbl>, .diff <dbl>,
#   .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>
\end{verbatim}

@@

The paper introduces a data pipeline comprising nine modules designed
for the construction and analysis of indexes within the tidy framework.
The pipeline offers a modular workflow to compute indexes with different
parameterizations, to test minor changes to the original index
definition, and to quantify uncertainties. The framework proposed in the
paper is universal to index across diverse domains. Examples are
provided, including the drought indexes (SPI and SPEI) and the Global
Gender Gap Index (GGGI), to demonstrate the index calculation with
different time scales and distributions, to illustrate how slight
adjustment of linear combination weights impact the index, and to
calculate confidence intervals on the index.

>>>>

\title{A Tidy Framework and Infrastructure to Systematically Assemble
Spatio-temporal Indexes from Multivariate Data}
\author{H. Sherry
Zhang$\textsuperscript{1,2}$~\orcidlink{0000-0002-7122-1463}, Dianne
Cook$\textsuperscript{2}$~\orcidlink{0000-0002-3813-7155}, Ursula
Laa$\textsuperscript{3}$~\orcidlink{0000-0002-0249-6439}, Nicolas
Langrené$\textsuperscript{4}$~\orcidlink{0000-0001-7601-4618}, Patricia
Menéndez$\textsuperscript{2,5}$~\orcidlink{0000-0003-0701-6315}}

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Science, University of Texas at Austin, Austin,
Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{3} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{5} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

@@

Now, we introduce the notation used for describing pipeline modules.
Consider a multivariate spatio-temporal process,
\[\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n \]
where:

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{DATA }\SpecialCharTok{|\textgreater{}} \FunctionTok{module1}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} \FunctionTok{module2}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} \FunctionTok{module3}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}\NormalTok{ ...}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
# A tibble: 128,576 x 18
  .idx  .dist id    month       ym  prcp  tmax  tmin  tavg  long   lat
  <chr> <chr> <chr> <dbl>    <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
1 spei  gev   ASN0~     6 1990 Jun   170  29.7  16.2  23.0  142. -15.5
2 spei  gev   ASN0~     7 1990 Jul   102  31.2  17.2  24.2  142. -15.5
3 spei  gev   ASN0~     8 1990 Aug     0  31.3  13.1  22.2  142. -15.5
4 spei  gev   ASN0~     9 1990 Sep     0  32.8  16.3  24.5  142. -15.5
5 spei  gev   ASN0~    10 1990 Oct     0  36.8  21.5  29.2  142. -15.5
# i 128,571 more rows
# i 7 more variables: name <chr>, .pet <dbl>, .diff <dbl>,
#   .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>
\end{verbatim}

@@

The paper introduces a data pipeline for constructing and analyzing
indexes within the tidy framework. The pipeline comprises nine modules
including temporal and spatial aggregation, variable transformation and
combination, distribution fitting, benchmark setting, and index
communication. The proposed pipeline is implemented as an R package,
\texttt{tidyindex}, available on CRAN.

One of the contributions of our work is its emphasis on aspects often
overlooked in index construction: uncertainty quantification and
sensitivity analysis. While indexes are widely constructed and used to
inform decision-making, such as on resource allocation, little effort is
made to justify choices made during indexes construction and to
experiment with alternatives. The pipeline proposed in the paper allows
for confidence interval calculation and sensitivity analysis to be
incorporated in the index construction process, for indexes across
various domains. These is illustrated through the three examples on
drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).

One future development of the work is to extend the tidyindex framework
to support index calculation on commonly used domain-specific data
formats, for example, NetCDF for climate indexes.
